---
title: "Our practice project"
author: "Mark kuria"
format:
  html:
    embed-resources: true
    number-sections: true
---

This Homework Project works with dataframes in the package `nycflights13` about flights, specifically a sample of domestic flights that departed from the three major New York City airports in 2013. 

```{r}
#| include: false

knitr::opts_chunk$set(echo = FALSE)
```
# Data overview 
```{r}
#| label: packages-data
#| message: false
#| echo: false

library(tidyverse)
library(patchwork)
library(nycflights13)
library(knitr)

```
## Package overview
-	The data package `nycflights13` package contains five tabular data frames. In total there is information about:   - `r nrow(airlines)` **airlines**
-	`r nrow(airports)` **airports**
-	`r nrow(planes)` **planes**
-	`r nrow(flights)` **flights**
The core dataframe `flights` has the following variables: 
`year`, `month`, `day`: **Date** of departure.  
`dep_time`, `arr_time`: Actual **departure** and **arrival** times (format HHMM or HMM), local time zone.  
`sched_dep_time`, `sched_arr_time`: **Scheduled departure** and **arrival** times (format HHMM or HMM), local time zone.  `dep_delay`, `arr_delay`: **Departure** and **arrival delays**, in minutes. Negative times represent early departures/arrivals.  
`carrier`: Two letter **carrier** abbreviation. See `airlines`-dataframe to get the full name.  

```{r}
#| label: flightsorigin
ggplot(data = flights, mapping = aes(x = origin)) + 
  geom_bar()
```
In the `flights` dataframe, the `EWR` airport has **`r nrow(filter(flights, 
origin=="EWR"))`** rows, the `JFK` has **`r nrow(filter(flights, origin=="JFK"))`** rows, and the LGA airport has **`r nrow(filter(flights, origin=="LGA"))`** rows, resulting in a total of **`r nrow(flights)`** flights.

```{r}
#| label: flightsdistance
ggplot(data = flights, mapping = aes(x = distance)) +   geom_histogram(binwidth = 50, fill = "orange", color = "black") +
  labs(
    title = "Distribution of Flight Distances",
    x = "Distance (miles)",     y = "Number of Flights"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(color = "gray80"),     panel.grid.minor = element_line(color = "gray90")
  )
```
```{r}
#| label: distributions # Scatter plot of seats
g1 <- ggplot(data = planes, mapping = aes(y = seats, x = 0)) + 
  geom_point(color = "orange") +
  labs(title = "Scatter Plot", x = "", y = "Seats") +   theme_minimal()
# Boxplot of seats
g2 <- ggplot(data = planes, mapping = aes(y = seats)) +   geom_boxplot(fill = "orange", color = "black") +   labs(title = "Boxplot", y = "Seats", x = "") +   theme_minimal()
# Histogram of seats
g3 <- ggplot(data = planes, mapping = aes(y = seats)) +   geom_histogram(binwidth = 10, fill = "orange", color = "black") +   labs(title = "Histogram", y = "Number of Planes", x = "Seats") +   theme_minimal()
# Combine the three plots
g1 + g2 + g3
```
The scatter plot (g1) shows each individual observation of seats in planes.  
The boxplot (g2) summarizes the distribution with median, quartiles, and outliers.  
The histogram (g3) shows the overall frequency distribution of seats. Each visualization has its advantages depending on what we want to see.
```{r}
#| label: engine-seats
ggplot(data = planes, mapping = aes(x = engine, y = seats)) +
  geom_boxplot(fill = "orange", color = "black") +
  labs(
    title = "Seats by Engine Type",
    x = "Engine Type",     y = "Seats"
  ) +   theme_minimal()
```
The boxplot shows how the number of seats varies by engine type.  
We can see which engine types typically have more or fewer seats and how spread out the seating capacities are.

```{r}
#| label: delays
#| message: false
#| warning: false

library(dplyr)

# Take a random sample of 5,000 flights
sample_flights <- sample_n(flights, 5000)

# Scatter plot of departure vs arrival delays with different colors for airports
ggplot(data = sample_flights, mapping = aes(x = dep_delay, y = arr_delay, color = origin)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Departure vs Arrival Delay",
    x = "Departure Delay (minutes)",
    y = "Arrival Delay (minutes)",
    color = "Origin Airport"
  ) +
  theme_minimal()

```

```{r}
#| label: airportlocations
ggplot(data = airports, mapping = aes(x = lon, y = lat, color =
tzone)) +
geom_point(alpha = 0.7) + coord_quickmap () +
labs (
title = "Airports by Longitude/Latitude and Time Zone",
x = "Longitude", y = "Latitude", color = "Time zone"
)
```
# Data Wrangling
```{r}
#| label: flightsaveragespeed
#| message: false
#| warning: false

flights |>
  filter(!is.na(air_time) & air_time > 0) |>
  mutate(speed = distance / (air_time / 60) * 1.60934) |>
  ggplot(aes(x = speed)) +
  geom_histogram(binwidth = 50) +
  labs(
    title = "Distribution of Flight Speeds",
    x = "Speed (km/h)",
    y = "Number of Flights"
  ) +
  theme_minimal()

```

```{r}
#| label: filtering library(knitr)
flights_filtered <- flights |>   filter(arr_delay >= 120)
flights_filtered |>
  select(year, month, day, dep_time, arr_time, carrier, dest, arr_delay) |>
  head(10) |>
  kable(caption = "Flights with arrival delay of 2 or more hours")
```

```{r}
#| label: summarizing
#| message: false
#| warning: false

nycflights13::flights |>
  summarize(
    `Average Departure Delay (minutes)` = round(mean(dep_delay, na.rm = TRUE), 0),
    .by = origin
  ) |>
  knitr::kable(caption = "Average Departure Delay per NYC Airport")
```
# Data Audit
```{r}
#| label: strangeairportlocations
#| message: false
#| warning: false

library(dplyr)
library(ggplot2)
library(knitr)

# Filter and arrange suspect airports (positive longitude)
suspects <- airports |>
  filter(lon > 0) |>
  select(faa, name, lat, lon, tzone) |>
  arrange(lon)

# Display a neat table
suspects |>
  kable(caption = "Airports with positive longitude (audit targets)")

# Plot suspect airports
ggplot(suspects, aes(lon, lat, label = faa)) +
  geom_point(size = 3, color = "orange") +
  geom_text(nudge_y = 1, check_overlap = TRUE) +
  coord_quickmap() +
  labs(
    title = "Suspect airport locations (positive longitude)",
    x = "Longitude",
    y = "Latitude"
  )

```

```{r}
#| label: flights-per-airline
#| message: false
#| warning: false

library(dplyr)
library(ggplot2)
library(forcats)

flights_count <- nycflights13::flights |>
  count(carrier, name = "n") |>
  left_join(nycflights13::airlines, by = "carrier") |>
  arrange(desc(n))

ggplot(flights_count, aes(x = n, y = fct_reorder(name, n))) +
  geom_col() +
  labs(
    title = "Number of Flights per Airline",
    x = "Number of Flights",
    y = "Airline"
  ) +
  theme_minimal()

```

```{r}
#| label: mean-dep-delay
#| message: false
#| warning: false

library(dplyr)
library(ggplot2)
library(forcats)

# Average departure delay per airline
delay_avg <- nycflights13::flights |>
  group_by(carrier) |>
  summarize(mean_dep_delay = mean(dep_delay, na.rm = TRUE), .groups = "drop") |>
  left_join(nycflights13::airlines, by = "carrier")

# Plot horizontal bar chart
ggplot(delay_avg, aes(x = mean_dep_delay, y = fct_reorder(name, mean_dep_delay))) +
  geom_col() +
  labs(
    title = "Mean Departure Delay per Airline",
    x = "Mean Departure Delay (minutes)",
    y = "Airline"
  ) +
  theme_minimal()

```


```{r}
#| label: side-by-side-delays
#| message: false
#| warning: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)

# Calculate mean departure and arrival delay per airline
delay_side <- nycflights13::flights |>
  group_by(carrier) |>
  summarize(
    mean_dep_delay = mean(dep_delay, na.rm = TRUE),
    mean_arr_delay = mean(arr_delay, na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(nycflights13::airlines, by = "carrier") |>
  pivot_longer(
    cols = c(mean_dep_delay, mean_arr_delay),
    names_to = "delay_type",
    values_to = "mean_delay"
  )

# Plot side-by-side bar chart
ggplot(delay_side, aes(x = mean_delay, y = fct_reorder(name, mean_delay), fill = delay_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(
    values = c(mean_dep_delay = "orange", mean_arr_delay = "blue"),
    labels = c(mean_dep_delay = "Departure Delay", mean_arr_delay = "Arrival Delay")
  ) +
  labs(
    title = "Mean Departure vs Arrival Delay by Airline",
    x = "Mean delay (minutes)",
    y = "Airline",
    fill = "Delay type"
  ) +
  theme_minimal()

```
**Note:** Some delays are negative because they represent flights that departed or arrived earlier than scheduled.  
-	Positive values → late  
-	Zero → on time  - Negative values → early
```{r}
#| label: planes-by-year
#| message: false
#| warning: false

library(dplyr)
library(ggplot2)

planes_renamed <- nycflights13::planes |>
  rename(manufacture_year = year)

planes_data <- nycflights13::flights |>
  left_join(planes_renamed, by = "tailnum") |>
  left_join(nycflights13::airports, by = c("origin" = "faa"))

planes_count <- planes_data |>
  filter(!is.na(manufacture_year)) |>
  group_by(name, manufacture_year) |>
  summarize(num_flights = n(), .groups = "drop")

ggplot(planes_count, aes(x = manufacture_year, y = num_flights)) +
  geom_col() +
  facet_wrap(~ name) +
  labs(
    title = "Number of Departing Planes by Year of Manufacture",
    x = "Year Manufactured",
    y = "Number of Flights"
  ) +
  scale_x_continuous(limits = c(1950, 2020)) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r}
#| label: weekly-precipitation
#| message: false
#| warning: false

library(dplyr)
library(lubridate)
library(ggplot2)

airports_lookup <- nycflights13::airports |>
  select(faa, name) |>
  rename(origin_name = name)

weekly_precip <- nycflights13::weather |>
  mutate(
    date = as.Date(paste(year, month, day, sep = "-")),
    week = lubridate::week(date)
  ) |>
  left_join(airports_lookup, by = c("origin" = "faa")) |>
  group_by(origin_name, week) |>
  summarize(total_precip = sum(precip, na.rm = TRUE), .groups = "drop")

ggplot(weekly_precip, aes(x = week, y = total_precip, color = origin_name)) +
  geom_line(linewidth = 1.2) +
  labs(
    title = "Total Weekly Precipitation at NYC Airports (2013)",
    x = "Week of Year",
    y = "Total Precipitation (inches)",
    color = "Airport"
  ) +
  theme_minimal()

```

```{r}
#| label: most-flights
#| message: false
#| warning: false

library(dplyr)
library(ggplot2)

flights_per_airline <- nycflights13::flights |>
  left_join(nycflights13::airlines, by = "carrier") |>
  group_by(name) |>
  summarize(total_flights = n(), .groups = "drop")

ggplot(flights_per_airline, aes(x = reorder(name, total_flights), y = total_flights)) +
  geom_col(fill = "orange", color = "black", linewidth = 0.2) +
  coord_flip() +
  labs(
    title = "Number of Flights per Airline (2013)",
    x = "Airline",
    y = "Total Flights"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(face = "bold")
  )

```
 
 ## 7.Do longer flights tend to have more arrival delay?
```{r}
#| label: distance-vs-arrival-delay
#| message: false
#| warning: false

library(dplyr)
library(ggplot2)

# Sample flights to keep the plot readable
sample_flights <- nycflights13::flights |>
  filter(!is.na(arr_delay), !is.na(distance)) |>
  sample_n(5000)

ggplot(sample_flights, aes(x = distance, y = arr_delay)) +
  geom_point(alpha = 0.4, color = "orange") +
  geom_smooth(method = "lm", color = "black", se = FALSE) +
  labs(
    title = "Relationship Between Flight Distance and Arrival Delay",
    x = "Flight Distance (miles)",
    y = "Arrival Delay (minutes)"
  ) +
  theme_minimal()
```

**Interpretation:**  
The scatter plot shows a wide spread of arrival delays across all flight distances. While longer flights do not always have larger arrival delays, the trend line suggests a slight positive relationship, meaning that longer flights tend to have somewhat higher arrival delays on average. However, the relationship is weak, indicating that factors other than distance likely play a larger role in arrival delays.
